\chapter{Introduction}

\section{Motivation}
Every year, dozens of new programming languages are proposed and each time this
happens,  the inventors of the new language have to answer the same question: 
There are hundreds of programming languages already, why do we need yet another one?  
Of course, the answer is always the same.  It consists of an economical argument, a
theological argument and a practical argument. For the
convenience of the reader, let us brievely review these arguments as they read for \setlx.  
\begin{enumerate}
\item Nothing less than the prosperity and welfare of the entire universe is at
      stake and \setlx\ provides the means to save it.
\item Programming in \setlx\ is the only way to guarantee redemption from 
      the eternal hell fire that awaits those not programming in \setlx.
\item Programming in \setlx\ is fun!
\end{enumerate}
The economical argument has already been discussed at length by Adams \cite{adams:1980},
therefore we don't have to repeat it here.   We deeply
regret the fact that the philosophical background of the average computer scientist does not permit
them to follow advanced theological discussions.  Therefore, we have to refrain from giving a
detailed proof of the second claim.  Nevertheless, we hope the examples given in this
tutorial will convince the reader of the truth of the third claim.
  One of the reasons for this is that \setlx\ programs are both very concise and
readable.  This often makes it possible to fit the implementation of complex algorithms in \setlx\ on a
single slide because the \setlx\ program is no longer than the pseudocode that is usually used to
present algorithms in lectures.  The benefit of this is that instead of pseudocode, students have a
running program that they can test and modify. This and the conciseness of \textsc{SetlX} programs was one of
the reasons for the first author to adopt \setlx\ as a programming language in his various lectures on computer 
science and mathematics: It is often feasible to write a complete \setlx\ program in a few lines 
onto the blackboard, since \setlx\ programs are nearly as compact as mathematical formul\ae.


\setlx\ is well suited to implement complex algorithms. This is achieved because SetlX
provides a number of sophisticated builtin data types that enable the user to code at a very high
abstraction level.
These data types are sets, lists, first-order terms, and functions.  
As sets are implemented as ordered binary trees, sets of pairs can represent
\href{https://en.wikipedia.org/wiki/Map_(mathematics)}{mappings} as well as
\href{https://en.wikipedia.org/wiki/Priority_queue}{priority queues}.  This enables very neat implementations
of a number of graph theoretical algorithms.  As \textsc{SetlX} is implemented in \textsl{Java} and
the \emph{Java virtual machine} is quite fast, various tests have shown that \textsc{SetlX} beats
\textsl{Python} in terms of performance. 


The purpose of this tutorial is to introduce the most important features of \setlx\ and to
show, how the use of the above mentioned data types leads to programs that are both shorter
and clearer than the corresponding programs in other programming languages.  This was the
prime motivation of the first author to develop \setlx:  It turns out that \setlx\ is very
convenient as a tool to present algorithms at a high abstraction level in a class room. 
Furthermore, \setlx\ makes the abstract concepts of set theory tangible for students of
computer science.  

\section{Overview}
The remainder
of this tutorial is structured as follows:
\begin{enumerate}
\item In the second chapter, we discuss the data types available in \setlx.
\item The third chapter provides the control structures.
\item The fourth chapter deals with \emph{regular expressions}.
\item The fifth chapter discusses \emph{functional programming} and \emph{closures}.
\item The sixth chapter discusses the \texttt{try-}\texttt{catch} and \texttt{throw}
      mechanism and demonstrates the use of \emph{backtracking}.
\item The seventh chapter discusses how vectors and matrices can be used and demonstrates the linear
      algebra package that is provided by \textsc{SetlX}.
\item The eighth chapter discusses the statistical functions provided by \textsc{SetlX}.
\item The ninth chapter presents the functions that can be used for plotting.
\item Chapter number ten introduces classes and demonstrates that \textsc{SetlX}
      supports an object-oriented programming style.
\item The final chapter lists and explains all predefined functions.
\item The appendix discusses the graphical primitives provided by \textsc{SetlX}.
      These primitives can be used, for example, to animate algorithms.
\end{enumerate}
This tutorial is not meant as an introduction to programming.  It assumes that the reader
has had some preliminary exposure to programming and has already written a few programs in
either \texttt{C}, \textsl{Java}, or a similar language.

\section{Downloading}
The current distribution of  \setlx\ can be downloaded from 
\\[0.2cm]
\hspace*{1.3cm}
\href{http://randoom.org/Software/SetlX}{\texttt{http://randoom.org/Software/SetlX}}.
\\[0.2cm]
\setlx\ is
written in \textsl{Java} and is therefore supported on a number of different operating
systems.  Currently, \setlx\ is supported on \textsl{Linux}, \textsl{OS X},
\textsl{Microsoft Windows}, and
\href{https://play.google.com/store/apps/details?id=org.randoom.setlxUI.android}{\textsl{Android}}.   
The websites given above explains how to install the language on various platforms.  
The distribution contains the \textsl{Java} code and a development guide that gives an
overview of the implementation. For those
interested in peeking under the hood, the source code of \textsc{SetlX}
is available at
\\[0.2cm]
\hspace*{1.3cm}
\href{https://github.com/herrmanntom/setlX}{\texttt{https://github.com/herrmanntom/setlX}}.



\subsection*{Disclaimer}
The development of \setlx\ is an ongoing project.  Therefore some of the material presented in
this tutorial might be out of date, while certain aspects of the language won't be
covered.  The current version of this tutorial is not intended to be a reference manual.
The idea is rather to provide the reader with an introduction that is sufficient to get started.

\subsection*{Encouragement}
The authors would be grateful for any kind of feedback.  They can be contacted via
email as follows:
\begin{tabbing}
\qquad \= Karl Stroetmann: \qquad \= \href{mailto:karl.stroetmann@dhbw-mannheim.de}{\texttt{karl.stroetmann@dhbw-mannheim.de}} \\[0.2cm]
       \> Tom Herrmann:           \> \href{mailto:setlx@randoom.org}{\texttt{setlx@randoom.org}}
\end{tabbing}


%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "tutorial"
%%% End: 
