\def\myDocumentTypeArticle{-}
\InputIfFileExists{\jobname_header}{}{\input{config/header_V001}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% document specific %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% define the author, title etc
\globalTitle{Development Guide}
\globalSubject{setlX \input{version}}
\globalAuthor{Herrmann, Tom}
\globalDate{\today}
\globalKeywords{readme}

% frequently used configuration options
\setboolean{linkHighlighting}{true}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% setup %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\InputIfFileExists{\jobname_configuration}{}{\input{config/configuration_V001}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% document text %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}
\begin{titlepage}
\maketitle
\vfill
\tableofcontents
\end{titlepage}

\section{Overview}

This is the development guide for \setlX, an interpreter for the \SetlX{} programming-language.

\section{Software Requirements}

The following software was used when developing the \setlX{} interpreter:

\begin{itemize}
	\item OpenJDK 1.6.0\_22, compatible to Oracles JDK 1.6 (aka version 6)
	\item Antlr 3.3, which is shipped as `.jar' with the interpreter
	\item pdfTeX, latex and various latex packages
	\item GNU Make 3.82
	\item Zip 3.0
\end{itemize}

\section{Distributions}

Three different `distributions' can be automatically created (on an Unix-like OS):

\begin{enumerate}
	\item A binary only distribution, which should work on all Unix-like OS and Windows without the need of the full Java JDK. Any Java Runtime (JRE) installation which is newer or the same major version as the JDK which created the distribution should work. This distribution does not include any \SetlX{} code.

	\item A source distribution, which must be rebuild by the user, which is automatically attempted on first launch. A Java JDK has to be installed for this to work. The version of the users JDK can be lower or even incompatible to the JDK which created the distribution. This distribution also does not include \SetlX{} code, except a simple program to test the interpreter (see `make test' in the manual).

	\item The development kit, which includes everything used in the development of the interpreter itself. This distribution includes various \SetlX{} example programs and the grammar of the interpreter in both `EBNF' and `pure' form.
\end{enumerate}

\subsection{Building distributable zip-files}

The automated distribution creation script can be launched by executing

\begin{lstlisting}
./createDistributions
\end{lstlisting}

Note that this process will `clean' the source (e.g. remove all automatically created files), rebuild the interpreter and all documentation and clean the source again.

\end{document}
