\input{config/header}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% document specific %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% define the author, title etc
\globalTitle{Interpreter Manual}
\globalSubject{setlX \input{version}}
\globalAuthor{Herrmann, Tom}
\globalDate{\today}
\globalKeywords{readme}

% frequently used configuration options
\setboolean{linkHighlighting}{true}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% setup %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{config/configuration}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% document text %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}
\begin{titlepage}
\maketitle

\vfill

\begin{center}
\Large
``Computers make very fast, very accurate mistakes.''
\end{center}

\vfill
\tableofcontents
\end{titlepage}

\section{Overview}

This is the manual of \setlX, an interpreter for the \SetlX{} (\underline{set} \underline{l}anguage e\underline{x}tended) programming-language.

The \SetlX{} language was mostly designed by Prof. Dr. Karl Stroetmann and is an evolution of \Setl{} by Jack Schwartz. It was specifically conceived to make the unique features of \Setl{} more accessible to today's students of set theories.

The \setlX{} interpreter, which currently is the \SetlX{} reference implementation, was implemented by Tom Herrmann. You may send bug-reports and\slash{}or questions about the \setlX{} interpreter via e-mail to \mailto{setlx@randoom.org}.


\section{Installation}

To comfortably use \setlX{} it may be installed into the search-path. After this is done, it may be launched just like any other command-line program on your system.

\subsection{Unix-like OS (Linux, MacOS X, etc.)}\label{Unix}

\begin{enumerate}
	\item Copy the `\command{setlX.jar}'-file into some folder, which is accessible by all users, who should be able to execute \setlX{} (e.g. `\command{/usr/local/setlX/}').\\
	You may also copy it into the home directory of some user, when he is the only \setlX-user on the system (e.g. `\command{/home/<user>/programs/setlX/}').
	\item Open the `\command{setlX}'-script in your favorite editor (either `\command{vi}' or `\command{emacs}' of course).
	\item Change the contents of the `\command{setlXJarLocation}'-variable to the full path of the copied jar-file, including the filename itself, e.g.
\begin{lstlisting}[frame=none,numbers=none]
setlXJarLocation="/usr/local/setlX/setlX.jar"
\end{lstlisting}
	\item Save and close the file.
	\item Copy the modified `\command{setlX}'-script into some folder, which is in the search-path of all users which should be able to execute \setlX{} (e.g. `\command{/usr/local/bin}').\\
	When only used by one user, you may copy it into the `bin' directory in his home instead (e.g. `\command{/home/<user>/bin}').
	\item Make the copied script executable, e.g.
\begin{lstlisting}[frame=none,numbers=none]
chmod +x /usr/local/bin/setlX
\end{lstlisting}
\end{enumerate}

\subsection{Microsoft Windows}

\begin{center}
\large
``Here's a nickel kid. Go buy yourself a real computer.''\footnote{In reference to \url{http://dilbert.com/1995-06-24}}
\end{center}

\begin{enumerate}
	\item Copy the `\command{setlX.jar}' and `\command{setlX.cmd}' files into some folder, which is accessible by all users, who should be able to execute \setlX{} (e.g. `\command{C:\textbackslash{}Progam Files\textbackslash{}setlX\textbackslash{}}').
	\item Open the copied `\command{setlX.cmd}'-script in your favorite text editor.
	\item Change the contents of the `\command{setlXJarLocation}'-variable to the full path of the jar-file, including the filename itself, e.g.
\begin{lstlisting}[frame=none,numbers=none]
set setlXJarLocation=C:\Progam Files\setlX\setlX.jar
\end{lstlisting}
	\item Save and close the file.
	\item Add the folder you placed both files into to the search-path (requires Administrator privileges):
	\begin{enumerate}
		\item Press \command{[Win]+[Pause]} key combination to open system preferences.
		\item 
		\begin{itemize}
			\item On Windows Vista or newer:\\Click on `\command{Advanced system settings}' on the left pane, which opens a new window.
			\item On Windows XP or older:\\Click the `\command{Advanced}' tab in the current window.
		\end{itemize}
		\item Click on `\command{Environment Variables}' on the bottom, which opens another window.
		\item Select the `\command{Path}' variable in the `\command{System variables}' section and click on `\command{Edit...}', which opens jet another window.
		\item Do not alter the existing content of the `\command{Variable value}' field, but add a semicolon (`\command{;}'), followed by the full path to the folder where you installed setlX, to its very end. Using the previous example path, the new value should look like:
\begin{lstlisting}[frame=none,numbers=none]
<previous value>;C:\Progam Files\setlX
\end{lstlisting}
		\item Accept all changes by clicking `\command{OK}' in all windows you opened before.
		\item End you current session and login into Windows again.
	\end{enumerate}
\end{enumerate}

\newpage

\section{Usage}

The \setlX{} interpreter has two basic modes of operation:

\begin{itemize}
	\item file execution mode, which can be launched with
\begin{lstlisting}[frame=none,numbers=none]
setlX <path>/<name>.stlx
\end{lstlisting}
	\item interactive mode, which can be launched with
\begin{lstlisting}[frame=none,numbers=none]
setlX
\end{lstlisting}
\end{itemize}

The `file execution mode' will be started if one or more paths to code files are supplied as parameters for this program. These files will then be parsed and executed in the order in which they are listed as parameters. When executing multiple files, they share the outermost scope.

When no paths are supplied, the `interactive mode' will be started.\\
Note that input will be executed after entering an additional empty line. Hereby multiple lines of input can be parsed and evaluated. Executing the `\command{exit;}' statement will terminate the interpreter.

\nonSubsection{Additional parameters}

In addition to file-paths a number of options can be used when running \setlX{}:

\begin{itemize}
	\param{--dump <file-name>}
	      {writes loaded code into a file (has no effect in interactive mode)}
	\param{--help}
	      {displays some helpful information}
	\param{--noExecution}
	      {load and check code for syntax errors, but do not execute it}
	\param{--predictableRandom}
	      {always returns the same pseudo random sequence of numbers from the internal random number generator (for debugging)}
	\param{--real32\\
	       --real64\\
	       --real128\\
	       --real256}
           {sets the width of the real-type in bits (64 is the sane default)}
	\param{--verbose}
           {display the parsed program before executing it (has no effect in interactive mode)}
	\param{--version}
           {displays the interpreter version and terminates}
\end{itemize}

\section{Building from source}

Optionally, when the interpreter source is available, the interpreter can be (re)build by executing

\begin{lstlisting}[frame=none,numbers=none]
make
\end{lstlisting}

in the `\command{src}' folder.

A self-contained `.jar' file will be created, which can be launched on all Java-Platforms without additional scripts, jars or environment-variables by executing

\begin{lstlisting}[frame=none,numbers=none]
java -jar setlX.jar <path>/<name>.stlx
\end{lstlisting}

or

\begin{lstlisting}[frame=none,numbers=none]
java -jar setlX.jar
\end{lstlisting}

However, the \command{setlX} and \command{setlX.cmd} scripts can also be used to launch the jar file created by this process.

\subsection{Building for Java 1.5 (esp. a 1st Generation Intel Macintosh)}\label{MACj5}

To create \setlX{} for a machine, which is limited to Java 1.5, you need another machine running both Java 1.5 and Java 1.6.

The main example for such situation is a 1st generation Apple Macintosh with Intel CPU running Mac OS X 10.5 or earlier. These machines are artificially limited by Apple to Java 1.5 only. It is recommended that you try installing an unofficial ``Soylatte''\footnote{\url{http://landonf.bikemonkey.org/static/soylatte/}} port of Java 1.6 for these machines, or upgrade to Mac OS X 10.6 or higher. You should only attempt to rebuild \setlX{} for Java 1.5 if you can not, or do not want to install this port.

On the second machine running Java 1.5 and 1.6, do the following:

\begin{enumerate}
	\item Activate Java 1.6.\\
		On a Mac this can be done using:
\begin{lstlisting}[frame=none,numbers=none]
/Applications/Utilities/Java Preferences
\end{lstlisting}
	\item Switch to the sub-directory src and execute the following command:
\begin{lstlisting}[frame=none,numbers=none]
make java-src/grammar/SetlXgrammar*.java
\end{lstlisting}
	\item Activate Java 1.5.\\
		On a Mac this can again be done using:
\begin{lstlisting}[frame=none,numbers=none]
/Applications/Utilities/Java Preferences
\end{lstlisting}
	\item Switch to the sub-directory src again and execute the following command:
\begin{lstlisting}[frame=none,numbers=none]
make
\end{lstlisting}
\end{enumerate}
This will create the file `\command{setlX.jar}', which can be installed on the machine running Java 1.5 as stated in section \ref{Unix}.

\section{System Requirements}

To run \setlX{} a Java runtime (JRE) is required, which is compatible to Java version 1.6 (aka version 6) or higher. Running it with Version 1.5 (aka version 5) is largely untested, but should work. However recompilation from source is required and somewhat problematic (see section \ref{MACj5}).

When building from source, the corresponding Java Development Kit (JDK) has to be present as well.

The hardware requirements are highly dependent on the executed \SetlX{} program. For optimal performance in most situations at least 512 MB of free main-memory should be available.

\section{Limitations}

\subsection{Stack Overflow \slash{} Out of Memory Error}

Due to its implementation and execution by the Java Virtual Machine (JVM), a `stack overflow' error may be encountered when executing highly recursive \SetlX{} programs. An `out of memory error' might also be encountered when using very large or very many sets and\slash{}or lists.

The JVM is unable to dynamically increase its internal stack size at all and the heap space only in very limited bounds, no matter how much free memory is available.

To work around these problems, edit the used launching script (`\command{setlX}' or `\command{setlX.cmd}') with a text editor and follow the directions given at the top of the script.

%\section{Known Bugs}
%The following bugs are known to be present in this version of the interpreter:
%
%\begin{itemize}
%	\item ...
%\end{itemize}


\section{Disclaimer}
You may distribute \setlX{}, to anyone and for any purpose, and make changes to it, even without permission from the author or notice to the author, as long as you do not remove any copyright information, and also comply with the antlr license.

This program uses the antlr parser generator in version 3.4. As per its license antlr is not guaranteed to work and might even destroy all life on this planet. The full text of this license is shown in figure \ref{lblAntlr}. Note however, that it only applies to the antlr components and NOT to the rest of the interpreter.

\includeListing[e]{../src/antlr/antlr_LICENSE.txt}{}{lblAntlr}{}{License used by antlr v3.4}


\end{document}
